Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-01-09 src/tex/ms.tex
[showyourwork] overleaf sync

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{Chempy} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percantage error of \variable{output/ape_NN.txt}, which is far below typical observational errors and even smaller away from the extremes of parameter space.
Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy}{https://github.com/TobiBu/sbi-chempy}. \textcolor{red}{adapt to point to final folder on the repo. and maybe add zenodo link for the nueral network weights...}

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_NN.png}
     \vspace{-.5cm}
     \caption{Absolute percentage error (APE) of the NN emulator for the \texttt{CHEMPY} simulator.}
     \label{fig:ape_NN}
     \script{train_torch_chempy.py}
\end{figure}

@@

\paragraph{Simulation-based inference}
In a nutshell, SBI \citep[e.g.][]{Cranmer2020,Papamakarios:2021,Gloeckler2024AllinoneSI} -- also called likelihood free inference within a  Bayesian inference framework --  works as follows: given an assumed generative model $\mathcal{M}$ of parameters $\theta$ (in our case a galactic chemical enrichment model) and a set of simulated observations of individual stellar abundances $\Vec{X}$ from that model, we train a mapping between the two to estimate the posterior distribution $p$($\theta|\Vec{X}$, $\mathcal{M}$) of the generative model parameters $\theta$ that reproduce the simulated observations $\Vec{X}$. Once this mapping is trained, we can apply it to real observations of stellar abundances $\Vec{X_R}$ to infer $p$($\theta|\Vec{X_R}$, $\mathcal{M}$)(see Fig. \ref{fig:flowchart}).
For this we use a Neural Posterior Estimator (NPE) \cite{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Masked Autoregressive Flow model (MAF) \cite{papamakarios2018maskedautoregressiveflowdensity} containing 5 transformer blocks with 50 hidden features each.

>>>>

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{Chempy} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percantage error of \variable{output/ape_NN.txt}, which is far below typical observational errors and even smaller away from the extremes of parameter space.
Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14507134}{https://zenodo.org/records/14507134}.

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_NN.png}
     \vspace{-.5cm}
     \caption{Absolute percentage error of the NN emulator for the \texttt{CHEMPY} simulator. The orange histogram shows the full distribution of percentage errors with the vertical dashed line indicating the median and the vertical dotted lines indicating the first and third quartile. The box plot on the top of the plot extends from the first quartile to the third quartile of the data, with a line at the median. The whiskers extend from the box to the farthest data point lying within $1.5\times$ the inter-quartile range from the box. The NN predicts abundances with an absolute percentage error far below typical observational errors.}
     \label{fig:ape_NN}
     \script{train_torch_chempy.py}
\end{figure}

@@

\paragraph{Simulation-based inference}
In a nutshell, SBI \citep[e.g.][]{Cranmer2020,Papamakarios:2021,Gloeckler2024AllinoneSI} -- also called likelihood free inference within a Bayesian inference framework -- works as follows: given an assumed generative model $\mathcal{M}$ of parameters $\theta$ (in our case a galactic chemical evolution model) and a set of simulated observations of individual stellar abundances $\Vec{X}$ from that model, we train a mapping between the two to estimate the posterior distribution $p$($\theta|\Vec{X}$, $\mathcal{M}$) of the generative model parameters $\theta$ that reproduce the simulated observations $\Vec{X}$. Once this mapping is trained, we can apply it to observations of stellar abundances $\Vec{X_R}$ to infer $p$($\theta|\Vec{X_R}$, $\mathcal{M}$)(see Fig. \ref{fig:flowchart}).

For this we use a Neural Posterior Estimator (NPE) \cite{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Masked Autoregressive Flow model (MAF) \cite{papamakarios2018maskedautoregressiveflowdensity} containing 10 hidden features and 1 transformation layer for the normalizing flow.

We train our NPE with $10^5$ data points simulated with the NN emulator described in the previous section until convergence. Training takes $\lesssim10$ minutes on an Intel(R) Xeon(R) Platinum 8468V with 192 CPUs.
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimick observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/posterior_ape.png}
     \vspace{-.5cm}
     \caption{Absolute percentage error of the neural posterior density estimate for a single star. Different coloured histograms show the full error distribution for all 6 parameters of interest with the median values highlighted by the vertical dashed lines. The box plots show again the first and third quartiles with the median represented by a vertical line and the whiskers extending from the box to the farthest data point lying within $1.5\times$ the inter-quartile range from the box. The global parameters of main interest for this work are shown by the light blue and red histogram.}
     \label{fig:ape_NN}
     \script{train_sbi.py}
\end{figure}

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate with 1000 data points (see Fig.~\ref{fig:posterior_samples} in the appendix). We then compare the posterior mean to the ground truth value. For a single star observation, our NPE has an APE of \variable{output/posterior_APE.txt} when looking at all 6 parameters of interest. If we restrict ourselves to only the global two parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$ our NPE achieves an APE of \variable{output/global_posterior_APE_NPE_C.txt} as we show in Fig.~\ref{fig:posterior_APE}. Looking at Fig.~\ref{fig:posterior_APE} we find that for an individual star the accuracy of the NPE is not particularly high. However, for the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$ we can boost the accuracy by combining the inference for many stars of the same galaxy.
